Liveness-Based Garbage Collection for Lazy Languages

### 2016-04-20 paper/paper.tex
fit in 10+bib pages

Experiments    with   a    single    generation   copying    collector
(Section~\ref{sec:experiments})   confirm  the   expected  performance
benefits.  Liveness-based collection results in an increase in garbage
reclaimed.  As  a consequence, there is  a reduction in the  number of
collections from 1.6X  to 23X and a decrease  in the minimum
memory requirement from  1X to 1198X.  As  an added benefit,
there is also a reduction in the overall execution time in 5 out of 12
benchmark programs.

@@

\begin{table*}[t!]
\caption{Statistics for liveness analysis and garbage collection}
\label{tab:exp-results}
\vskip -2mm
\centering
\include{data-liveness-analysis}
\vskip -5mm
\include{autotable}
\vskip -7mm
\end{table*}

%\subsection{Garbage collection for references on print stack}
The evaluation of  the top-level expression in a program  is driven by
a printing function~(Section
\ref{sec:semantics}).  
%% Printing of  atomic values is simple  and does
%% not  trigger a  GC.   In case  of  \CONS\ cells,  the
%% \CAR\ is  first printed followed  by \CDR.   In a lazy  language, both
%% \CAR\ and \CDR\  could be closures requiring evaluation,  and this may
%% trigger  more GCs.  When  this happens.   we have  to
%% consider any  references that  might be  on the  print stack  and copy
%% them.
We extend liveness-based  garbage  collection  to
this function.
\section{Experimental evaluation}
\label{sec:experiments}
Our  experimental  setup   consists  of  the  prototypes   of  (a)  an
interpreter  for our  language, (b)  a  liveness analyzer,  and (c)  a
single generation  copying garbage  collector.  The  garbage collector
can be configured  to work on the basis of  reachability (RGC mode) or
use liveness DFAs (LGC mode).
%
Our benchmark consists of programs taken from nofib~\cite{nofib} and
other repositories for functional programs~\cite{PLT-Scheme, gc_bench,
  huffman-sicp}.  We ran the experiments on 8 core
Intel\textsuperscript{\textregistered} Core\textsuperscript{TM}
i7-4770 3.40GHz CPU having 8192KB L2 cache, 16GB RAM, running 64 bit
Ubuntu 14.04.

@@

Table~\ref{tab:exp-results}(b) compares  GC statistics
for RGC and  LGC.  We report the number of  GC events,
average  number of  cells  reclaimed per  GC,  average
number of cells  touched per GC and the  total time to
perform all collections.   It is no surprise that the  number of cells
reclaimed per garbage  collection is higher and the  number of garbage
collections  lower  for  LGC.   The  cost of  LGC  is  higher  garbage
collection time, which increases the  overall execution time even with
reduced number of  collections.  However, the execution time  of LGC is
still comparable for most benchmarks (slowdown within 5X of
RGC in most cases) and better  for 3 benchmarks (2X speedup in the
best case).  Note that  \verb@gc_bench@~\cite{gc_bench} is a synthetic
benchmark that allocates complete binary
trees  of various  sizes that  are  never used  by the  program. As  a
result,  the benchmark  highly favours  LGC.  The  benchmark has  been
included  for  completeness,  and  we  do  not  consider  its  numbers
as being representative of real programs.

@@

Note that our proofs will be for a single round of minefield execution
i.e.,  the evaluation of  $(\mainpgm)$  to   its  WHNF  driven  by   the  printing  mechanism~(Section
\ref{sec:semantics}).  With  minor variations,  the proof will  also be
applicable for subsequent evaluations initiated by $\print$.
%% We show  next that the  minefield execution of a  call-less program
%% cannot go \bang.

\begin{lemma}
\label{lemma:call-less-cannot-go-bang}
Consider the minefield execution of  a program without function calls.
Such a program cannot enter the \bang\ state.
\end{lemma}
\begin{proof}
Consider a  state  $(\rho,  S,  \heap,  e, \sigma)$  in  the
minefield execution of a program.  We  show by induction on the number $n$
of $\rightsquigarrow$ steps  leading to  this state  that the next  transition cannot
enter a  \bang\ state.   When $n$  is 0,  the state  is $([\;]_\rho,\,
(\rho_\mathit{init},   \,  \ell_{\ans}   ,\,  (\print~\ans)):[\;]_{S},
[\;]_{\heap},\, e_{\mainpgm})$.   Since the call  to $GC(\ldots)$ in  this state
does  nothing,  we  just  have to  show  that  the  $\rightsquigarrow$
transition cannot  enter a  \bang\ state.  Since  our programs  are in
ANF, $e_{\mainpgm}$ can only be a $\LET$ expression.  A {\sc let} step
does not involve dereferencing, and thus cannot result in a \bang.

@@

>>>>

Experiments    with   a    single    generation   copying    collector
(Section~\ref{sec:experiments})   confirm  the   expected  performance
benefits.  Liveness-based collection results in an increase in garbage
reclaimed.  As  a consequence, there is  a reduction in the  number of
collections from 1.6X  to 23X and a decrease  in the minimum
memory requirement from  1X to 1198X.  As  an added benefit,
there is also a reduction in the overall execution time in 4 out of 12
benchmark programs.

@@

\begin{table*}[t!]
\caption{Statistics for liveness analysis and garbage collection}
\label{tab:exp-results}
\vskip -2mm
\centering
\include{data-liveness-analysis}
\vskip -5mm
{\color{Myblue}
\include{autotable}
}
\vskip -7mm
\end{table*}

%\subsection{Garbage collection for references on print stack}
The evaluation of  the top-level expression in a program  is driven by
a printing function~(Section
\ref{sec:semantics}).  
%% Printing of  atomic values is simple  and does
%% not  trigger a  GC.   In case  of  \CONS\ cells,  the
%% \CAR\ is  first printed followed  by \CDR.   In a lazy  language, both
%% \CAR\ and \CDR\  could be closures requiring evaluation,  and this may
%% trigger  more GCs.  When  this happens.   we have  to
%% consider any  references that  might be  on the  print stack  and copy
%% them.
We extend liveness-based  garbage  collection  to
this function.
\section{Experimental evaluation}
\label{sec:experiments}
{\color{Myblue}
Our  experimental  setup   consists  of  the  prototypes   of  (a)  an
interpreter  for our  language, (b)  a  liveness analyzer,  and (c)  a
single generation  copying garbage  collector.  The  garbage collector
can be configured  to work on the basis of  reachability (RGC mode) or
use liveness DFAs (LGC mode).}
%
Our benchmark consists of programs taken from nofib~\cite{nofib} and
other repositories for functional programs~\cite{PLT-Scheme, gc_bench,
  huffman-sicp}.  We ran the experiments on 8 core
Intel\textsuperscript{\textregistered} Core\textsuperscript{TM}
i7-4770 3.40GHz CPU having 8192KB L2 cache, 16GB RAM, running 64 bit
Ubuntu 14.04.

@@

Table~\ref{tab:exp-results}(b) compares  GC statistics
for RGC and  LGC.  We report the number of  GC events,
average  number of  cells  reclaimed per  GC,  average
number of cells  touched per GC and the  total time to
perform all collections.   It is no surprise that the  number of cells
reclaimed per garbage  collection is higher and the  number of garbage
collections  lower  for  LGC.   The  cost of  LGC  is  higher  garbage
collection time, which increases the  overall execution time even with
reduced number of  collections.  However, the execution time  of LGC is
still comparable for most benchmarks (slowdown within 5X of
RGC in most cases) and better  for 3 benchmarks (2X speedup in the
best case).  
%% Note that  \verb@gc_bench@~\cite{gc_bench} is a synthetic
%% benchmark that allocates complete binary
%% trees  of various  sizes that  are  never used  by the  program. As  a
%% result,  the benchmark  highly favours  LGC.  The  benchmark has  been
%% included  for  completeness,  and  we  do  not  consider  its  numbers
%% as being representative of real programs.

@@

Our proofs will be for a single round of minefield execution i.e., the
evaluation  of  $(\mainpgm)$  to  its  WHNF  driven  by  the  printing
mechanism~(Section \ref{sec:semantics}).   With minor  variations, the
proof will also be applicable  for subsequent evaluations initiated by
$\print$.
%% We show  next that the  minefield execution of a  call-less program
%% cannot go \bang.
%
\begin{lemma}
\label{lemma:call-less-cannot-go-bang}
Consider the minefield execution of  a program without function calls.
Such a program cannot enter the \bang\ state.
\end{lemma}
\begin{proof}
Consider a  state  $(\rho,  S,  \heap,  e, \sigma)$  in  the
minefield execution of a program.  We  show by induction on the number $n$
of $\rightsquigarrow$ steps  leading to  this state  that the next  transition cannot
enter a  \bang\ state.   When $n$  is 0,  the state  is $([\;]_\rho,\,
(\rho_\mathit{init},   \,  \ell_{\ans}   ,\,  (\print~\ans)):[\;]_{S},
[\;]_{\heap},\, e_{\mainpgm})$.   Since the call  to $GC(\ldots)$ in  this state
does  nothing,  we  just  have to  show  that  the  $\rightsquigarrow$
transition cannot  enter a  \bang\ state.  Since  our programs  are in
ANF, $e_{\mainpgm}$ can only be a $\LET$ expression.  A {\sc let} step
does not involve dereferencing, and thus cannot result in a \bang.

@@

\section*{Acknowledgement}
We would like to thank the anonymous referees for their valuable suggestions. We thank 
Evgeny Kapun for his suggestions regarding the undecidability proof.
